---
format:
  docx:
    reference-doc: "/Users/lamhine/Documents/manuscript-templates/manuscript-reference.docx"
    number-sections: false
    fig-dpi: 300
bibliography: references.bib
csl: "/Users/lamhine/Documents/manuscript-templates/american-medical-association.csl"
link-citations: false
link-bibliography: false
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "png"
    dpi: 300
---

```{r setup}
#| include: false

library(dplyr)
library(here)
library(flextable)

df_rates <- readRDS(here("01_data", "df_rates.rds"))
desc     <- readRDS(here("01_data", "desc_totals_50plus.rds"))

n_total  <- desc$overall
n_female <- desc$by_sex$n[desc$by_sex$sex == "Female"]
n_male   <- desc$by_sex$n[desc$by_sex$sex == "Male"]

fmt_n <- function(x) format(x, big.mark = ",")

rate <- function(grp, sx, age = "85+") {
  x <- df_rates |> filter(group == grp, sex == sx, age_output == age)
  formatC(x$estimate, digits = 1, format = "f")
}

main_word_count <- function(path = here("04_manuscript", "manuscript.qmd")) {
  lines <- readLines(path)
  start <- which(grepl("^# Introduction\\s*$", lines))[1]
  end_candidates <- which(grepl(">Acknowledgments</w:t>", lines, fixed = TRUE))
  end <- end_candidates[end_candidates > start][1]
  if (is.na(start) || is.na(end)) return(NA)
  lines <- lines[(start + 1):(end - 1)]

  lines <- lines[!grepl("^#", lines)]
  in_raw <- FALSE
  keep <- rep(TRUE, length(lines))
  for (i in seq_along(lines)) {
    if (grepl("^```\\{=openxml\\}", lines[i])) { in_raw <- TRUE; keep[i] <- FALSE }
    else if (grepl("^```\\s*$", lines[i]) && in_raw) { in_raw <- FALSE; keep[i] <- FALSE }
    else if (in_raw) keep[i] <- FALSE
  }
  text <- paste(lines[keep], collapse = " ")

  text <- gsub("\\[@[^\\]]+\\]", "", text)
  text <- gsub("`r [^`]+`", "X", text)
  text <- gsub("\\*\\*|\\*|__|_|`", "", text)
  text <- gsub("\\^[^\\s\\^]+\\^", "", text)
  text <- gsub("[\\\\(){}\\[\\]]", " ", text)

  words <- strsplit(trimws(text), "\\s+")[[1]]
  length(words[nzchar(words)])
}
```

::: {custom-style="Author"}
**Title:** Assault injuries among older Multiracial adults in California, 2005-2022

 

**Authors:** Tracy Lam-Hine, DrPH, MBA^1^; Emily F. Liu, MPH^2^; Caitlin Chan, SM^2^; Jennifer Ahern, PhD, MPH^2^

^1^ Stanford University School of Medicine, Department of Epidemiology and Population Health\
^2^ University of California, Berkeley, Division of Epidemiology, School of Public Health

 

**Corresponding author:**\
Tracy Lam-Hine, DrPH, MBA\
Stanford University School of Medicine, Department of Epidemiology and Population Health\
1701 Page Mill Road, Palo Alto, CA 94304\
lamhine\@stanford.edu

 

**Article type:** Research Letter

**Short running title:** Assault injuries among older Multiracial adults

**Word count (main text):** `r main_word_count()`

**Tables:** 1

**Figures:** 1

**Financial support:** This research received funding from the Eunice Kennedy Shriver National Institute of Child Health and Human Development and the National Institutes of Health Office of the Director (DP2HD080350). JA is a Chan Zuckerberg Biohub-San Francisco Investigator. TLH did not receive direct funding for this paper.

**Prior presentation:** Accepted for poster presentation at the Society for Epidemiologic Research Annual Meeting, 2026.
:::

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:br w:type="page"/>
  </w:r>
</w:p>
```
# Introduction

Violence against older adults is a growing public health concern. Between 2015 and 2022, U.S. emergency department visit rates for nonfatal assault among adults aged 60 years and older rose 31%, and homicide rates rose 26%, with an estimated annual economic cost of \$32.7 billion in 2022.[@peterson2024] Nationally, racial and ethnic disparities in nonfatal assault injury among older adults have been poorly characterized due to high levels of missing race and ethnicity data in the National Electronic Injury Surveillance System.[@logan2019] A recent analysis of assault injuries among women in California found that Multiracial women aged 65 years and older had the highest age-specific assault injury rates of any racial and ethnic group, exceeding Black women in this age range.[@liu2026] Whether this pattern extends to Multiracial men or varies across the older age range is unknown. We estimated rates of fatal and non-fatal assault among California adults aged 50 years and older from 2005 to 2022 – stratifying by age group, sex, and race and ethnicity – to characterize patterns among Multiracial older adults across the older age range and in both sexes.

# Methods

We combined 2005-2022 statewide hospital discharge data from the California Department of Health Care Access and Information with death records from the California Department of Public Health for residents aged 50 years and older. Using the data infrastructure and case definition described in Liu et al.,[@liu2026] we identified fatal assault injuries from death records using ICD-10 external cause codes and non-fatal assault injuries from hospital records using ICD-9-CM external cause codes for events before October 2015 and ICD-10-CM external cause and diagnostic codes for events from October 2015 onward. To avoid double counting, we linked ED visits resulting in admission to their inpatient records and removed in-hospital deaths from the hospital data. We excluded records with missing age or race and ethnicity (\<5%). We classified race and ethnicity into six mutually exclusive groups: Hispanic (any race), and non-Hispanic American Indian or Alaska Native (AIAN), Asian or Pacific Islander (API), Black, Multiracial, and White. We categorized age into 5-year intervals from 50-54 through 85+. We calculated sex- and age-specific assault injury rates per 100,000 person-years using intercensal (2005-2009) and postcensal (2010-2022) California population estimates as denominators. Exact 95% Poisson confidence intervals were estimated using the epitools package in R.[@aragon2004] Use of these data was approved by the California Health and Human Services Agency and the University of California, Berkeley Committee for the Protection of Human Subjects.

# Results

Between 2005 and 2022, we identified `r fmt_n(n_total)` assault injury events among California adults aged 50 years and older (`r fmt_n(n_female)` among females and `r fmt_n(n_male)` among males). Assault injury rates declined with age for females and males across all racial and ethnic groups (**Figure 1**). Across ages 50 to 64 years, Black adults had the highest assault injury rates. At ages 65 and older, Multiracial adults had the highest and most persistently elevated rates across groups.

Among adults aged 85 years and older, assault injury rates among Multiracial adults were `r rate("MULT", "Female")` per 100,000 person-years among females and `r rate("MULT", "Male")` per 100,000 person-years among males, compared with `r rate("BAA", "Female")` and `r rate("BAA", "Male")` among Black females and males, and `r rate("WHITE", "Female")` and `r rate("WHITE", "Male")` among White females and males, respectively (**Table 1**).

# Discussion

Assault injury rates among older Multiracial adults in California were elevated relative to other racial and ethnic groups and showed less age-related decline. These findings extend a recent analysis among California women by showing the same pattern in both sexes and across the older age range.[@liu2026] Multiracial older adults are routinely excluded from or aggregated within aging research, obscuring patterns of late-life risk that may differ from those of other groups.[@lam-hine2025]

Several mechanisms may contribute to this pattern. Multiracial adults experience elevated rates of social isolation,[@town2024] which is an established risk factor for elder mistreatment and late-life interpersonal violence.[@rosen2019] Multiracial individuals also frequently report identity invalidation and marginalization across multiple racial communities,[@franco2021] which may further weaken protective family and community ties as they age. These mechanisms warrant direct investigation in older Multiracial populations.

This analysis had limitations. Hospital and death records capture only assault injuries severe enough to result in medical attention or death and do not capture non-medically-attended events. These data also do not include information on perpetrator identity or relational context, and findings should not be interpreted as estimates of elder abuse incidence.

As the older adult population grows and diversifies, research disaggregating Multiracial older adults is needed to understand the conditions shaping late-life assault injury risk and to inform prevention efforts.

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:br w:type="page"/>
  </w:r>
</w:p>
```
```{=openxml}
<w:p>
  <w:pPr>
    <w:pStyle w:val="Heading1"/>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:t xml:space="preserve">Acknowledgments</w:t>
  </w:r>
</w:p>
```
::: {custom-style="Author"}
**Conflict of Interest Statement:** The authors declare no conflicts of interest.

**Role of Funder:** Funders had no involvement in the study's conceptualization, design, analysis, or interpretation of results.

**Access to Data:** EFL had full access to the underlying individual-level data used to generate the aggregated rates analyzed in this manuscript. TLH had full access to the aggregated data analyzed in this manuscript and takes responsibility for the integrity of the analysis and figures presented.

**Data Sharing Statement:** Hospital discharge data used for this analysis are available upon request from the California Department of Health Care Access and Information. Death record data are available upon request from the California Department of Public Health.
:::

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:br w:type="page"/>
  </w:r>
</w:p>
```
```{=openxml}
<w:p>
  <w:pPr>
    <w:pStyle w:val="Heading1"/>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:t xml:space="preserve">References</w:t>
  </w:r>
</w:p>
```
::: {#refs}
:::

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:br w:type="page"/>
  </w:r>
</w:p>
```
::: {custom-style="Table Caption"}
**Table 1.** Age-specific assault injury rates per 100,000 person-years (95% CI) by race and ethnicity and sex, California, 2005-2022.
:::

```{r tbl1}
readRDS(here("03_results", "tbl_rates.rds"))
```

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
  <w:r>
    <w:br w:type="page"/>
  </w:r>
</w:p>
```
::: {custom-style="Image Caption"}
**Figure 1.** Assault injury rates per 100,000 person-years by age group, race and ethnicity, and sex, California, 2005-2022.
:::

::: {custom-style="Compact"}
```{r fig1}
#| fig-width: 6.2
#| fig-height: 3.3
#| fig-align: left

readRDS(here("03_results", "fig_rates.rds"))
```
:::

::: {custom-style="Image Caption"}
Shaded bands represent 95% confidence intervals. Abbreviations: AIAN = American Indian or Alaska Native; API = Asian or Pacific Islander.
:::

```{=openxml}
<w:p>
  <w:pPr>
    <w:suppressLineNumbers/>
  </w:pPr>
</w:p>
```
